---
title: "DWARF PARAFAC"
author: "CBG"
date: "3/23/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load required packages, include=FALSE}
Packages <- c("staRdom","dplyr", "tidyr")
lapply(Packages, library, character.only = TRUE)
```

## DWARF Fluorescence
- first attempt to PARAFAC the fluorescence EEM 
- can we distinguish anthropogenic from natural DOM?
```{r Load and combine data}
#library(devtools)
#install_github("MatthiasPucher/staRdom")
#library(staRdom)
eem_list1 = eem_import_dir("C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/DWARF-PARAFAC/new/new2/new4")
#summary(eem_list2)
#sessionInfo()
```

- seems like eem_list2, 3, 4 and 5 is not blank corrected?
```{r View EEM data}
#eem_overview_plot(eem_list1, spp=9, contour = TRUE)
dwarf_sum <- summary(eem_list1)  

#following to find samples to keep
write.csv(dwarf_sum,"C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/DWARF_summary2.csv", row.names = FALSE)
```

- Exclude the following samples that do not belong to the DWARF project
```{r exclude samples that dont belong to the dwarf project}
#The noisy range below 250 nm excitation and above 580 nm emission can be removed 
info <- read.table("DWARF_sampleoverview3.txt", sep="\t", header=TRUE)
#Otava <- subset(info, Where=="Otava")
Ex_list <- subset(info, Where==c("DWTP Pisek", "Karhov"))

#Otava_samples <- as.list(Otava$DWARF_fluorescence)
Ex_samples <- as.list(Ex_list$DWARF_fluorescence)

#keep <- list("ex" = c(),
#                "em" = c(),
#                "sample" = Otava_samples)

exclude <- list("ex" = c(),
                "em" = c(),
                "sample" = Ex_samples)

exclude2 <- list("ex" = c(),
                "em" = c(),
                "sample" = c("v2021135", "v2021136","v2021137","v2021138","v2021139","v2021140","v2021141","v2021142","v2021143","v2021144","v2021145","v2021146","v2021147","v2021148","v2021149","v2021150","v2022117","v2022118","v2022119","v2022120","v2022121","v2022122","v2022123","v2022124","v2022125","v2022126","v2022127","v2022128","v2022129","v2022130","v2022153", "v2022154","v2022155","v2022156","v2022157","v2022158","v2022159","v2022160","v2022161","v2022164","v2022165","v20212837","v20212838","v20212839","v20212840","v20212841","v20212842","v20212843","v20212844","v20212845","v20212846","v20212847","v20213112","v20213113","v20213114","v20213142","v20213143","v20213144","v20213223","v20213224","v20213225","v20213226","v20213231","v20213232","v20213234","v20213276","v20213277","v20213278","v20213279","v20213280","v20213281","v20213371","v20213372","v20213373","v020213141NF","v020213142NF","v020213143NF","v020213144NF","v020213222NF","v020213223NF","v020213224NF","v020213225NF","v20213226NF","v20213233NF","v20213234NF","v20213275NF","v20213276NF","v20213277NF","v20213278NF","v20213279NF","v20213280NF","v20213281NF","v20213370NF","v20213371NF","v20213372NF","v20213373NF","v20213141NF","v20213142NF","v20213143NF","v20213144NF","v20213222NF","v20213223NF","v20213224NF","v20213225NF"))

#print(exclude)
# exclude outliers if neccessary. if so, restart analysis
eem_list_ex <- eem_exclude(eem_list1, exclude)
eem_list_ex2 <- eem_exclude(eem_list_ex, exclude2)
summary(eem_list_ex2)
length(eem_list_ex2)


eem_list2 <- eem_list_ex2 %>% eem_range(ex = c(250.000,Inf), em = c(300.000,581.000))
eem_overview_plot(eem_list2, spp=9, contour = TRUE)

dwarf_sum <- summary(eem_list2)  
write.csv(dwarf_sum,"C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/230329_DWARF_summaryX.csv", row.names = FALSE)
```

```{r First model attempt}
 #10) First attempt PARAFAC model
# minimum and maximum of numbers of components
# parameter settings from stardom guideline
cores <- detectCores(logical = FALSE)
dim_min <- 2
dim_max <- 4

nstart <- 25 # number of similar models from which best is chosen
maxit = 5000 # maximum number of iterations in PARAFAC analysis
ctol <- 10^-6 # tolerance in PARAFAC analysis

# 11) calculating PARAFAC models, one for each number of components
#using non-negative constraints
pf1 <- eem_parafac(eem_list2, comps = seq(dim_min,dim_max), normalise = FALSE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)

# 12) rescale B and C modes to a maximum fluorescence of 1 for each component
pf1n <- lapply(pf1, eempf_rescaleBC, newscale = "Fmax")
#Compare the different models in plot
print(eempf_compare(pf1n, contour = TRUE))
```


```{r First model checking: correlation between components}
# 13) check for correlation between the components
#The number indicates which model to look at
print(eempf_cortable(pf1n[[2]], normalisation=FALSE))
print(eempf_corplot(pf1n[[2]], progress = FALSE, normalisation = FALSE))

# 14) Normalise to reduce correlation between components
pf1x <- eem_parafac(eem_list2, comps = seq(dim_min,dim_max), normalise = TRUE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
# rescale B and C modes
pf1xn <- lapply(pf1x, eempf_rescaleBC, newscale = "Fmax")
print(eempf_compare(pf1xn, contour = TRUE, normalisation = FALSE))

#Set normalisation TRUE if you want to see the actual data
#Set normalisation FALSE if you want to see the normalised data
eempf_cortable(pf1xn[[2]], normalisation = FALSE)
eempf_corplot(pf1xn[[2]], progress = FALSE, normalisation = FALSE)
eempf_compare(pf1xn, contour = TRUE, normalisation=FALSE) 
```

```{r Look for outliers in leverages}
# 15) Find and exclude outlier leverages
cpl <- eempf_leverage(pf1xn[[2]])
# plot leverage (nice plot)
eempf_leverage_plot(cpl,qlabel=0.1)

# samples, excitation and emission wavelengths to exclude, makes sense after calculation of leverage
exclude1 <- list("ex" = c("255","250", "275", "280", "285"),
                "em" = c(),
                "sample" = c("v20222340", "v20221453", "v20221440")
)

# exclude outliers if neccessary. if so, restart analysis
eem_list_ex1 <- eem_exclude(eem_list2, exclude1)

# plot leverage, not so nice plot but interactive to select what to exclude
# saved in exclude, can be used to start over again with eem_list_ex <- eem_list %>% eem_exclude(exclude) above
exclude <- eempf_leverage_ident(cpl,qlabel=0.1)


# DID NOT MANAGE THE FOLLOWING. SOMETHING TO DO WITH SAMPLE NAMES?
eempf_residuals_plot(pf1xn[[2]], eem_list2, residuals_only = TRUE, 
                     select = c("v20222340", "v20221453", "v20221440"), spp=6, cores = cores, contour = TRUE)
```
- should the ex wavelengths with higher leverage be excluded?
- should the two samples really be excluded? No, not from looking at the leverages
```{r Create new model without outliers}
pf2 <- eem_parafac(eem_list_ex1, comps = seq(dim_min,dim_max), normalise = TRUE, maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
pf2n <- lapply(pf2, eempf_rescaleBC, newscale = "Fmax")

eempf_compare(pf2n, contour = TRUE, normalisation = FALSE)
eempf_leverage_plot(eempf_leverage(pf2n[[2]]),qlabel=0.1)



```
#EValuating firt attempt
- from model fit the model2 with 3 comps seems best
- the ex and em spectra are not 100% valid but might be ok?
- HIGH correlation between components

Not sure of two or three components
Make two different models
Interestingly, the third component looks identical to 1000Lakes assumed contamination
```{r PARAFAC increased}
# 16) ruN MODEL AGAIN WITH INCREASED
# minimum and maximum of numbers of components
nstart <- 50 # number of similar models from which best is chosen
maxit = 10000 # maximum number of iterations in PARAFAC analysis
ctol <- 10^-10 # tolerance in PARAFAC analysis

#here we specify the number of components and so there is only one model
pf3c <- eem_parafac(eem_list2, comps = 3, const = c("nonneg", "nonneg", "nonneg"), normalise = TRUE, maxit = maxit, nstart = nstart, output = "all", ctol = ctol, cores = cores)
pf3cn <- lapply(pf3c, eempf_rescaleBC, newscale = "Fmax")
eempf_compare(pf3cn, contour = TRUE)
```

```{r Model checking}
 # 17) Check the convergence behavior of the created models:
eempf_convergence(pf3cn[[1]])
eempf_compare(pf3cn, contour=TRUE)
eempf_leverage_plot(eempf_leverage(pf3cn[[1]]))
eempf_corplot(pf3cn[[1]], progress = FALSE)

#plot the resulting components and loadings
eempf_comp_load_plot(pf3cn[[1]], contour = TRUE)

# plot components in each sample, residual and whole sample
#WHen using this plot, perhaps choose only some samples at the time? Did not work now with all
#eempf_residuals_plot(pf3cn[[1]], eem_list2, cores = cores, contour = TRUE)

#calculate split_half analysis
sh <- splithalf(eem_list2, 3, normalise = TRUE, rand = FALSE, cores = cores, nstart = nstart, strictly_converging = TRUE, maxit = maxit, ctol = ctol)
splithalf_plot(sh)
sh %>% unlist(recursive = FALSE) %>% eempf_compare()

# 20) Tucker’s Congruency Coefficients is a value for the similarity of the splits
# 1 would be perfect similarity
tcc_sh_table <- splithalf_tcc(sh)
tcc_sh_table

sh_r <- splithalf(eem_list2, 3, normalise = TRUE, rand = TRUE, cores = cores, nstart = nstart, maxit = maxit, ctol = ctol)
splithalf_plot(sh_r)
tcc_sh_table2 <- splithalf_tcc(sh_r)
tcc_sh_table2

varimp <- eempf_varimp(pf3cn[[1]], eem_list2, cores = cores)
varimp
```

```{r Model Export}
#Export to OpenFluor
eempf_openfluor(pf3cn[[1]], file = "22_CBG_DWARF_3comps_2.txt")
#Export model for use
eempf_export(pf3cn[[1]], export = "C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/220401_PARAFAC_DWARF_3comp.txt", Fmax = FALSE)


#Make report
eempf_report(pf5n[[1]], export = "parafac_report.html", eem_list = eem_list_ex, shmodel = sh, performance = TRUE)
#eempf_report(pf3[[1]], export = "parafac_report.html")

#Reverse normalisation
p4 <- norm2A(pf5[[1]])

#Export model
#No difference in output from Fmax TRUE or FALSE for pf5m(1)
#p4 fra pf5n ga samma output som over
#Now they represent two different outputs, BUT they are relatively the same, the variation among the samples
eempf_export(pf_2cn[[1]], export = "C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/PARAFAC_DWARF_2comp.txt", Fmax = FALSE)
eempf_export(pf_3cn[[1]], export = "C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/PARAFAC_DWARF_3comp.txt", Fmax = FALSE)
#the fluorescence indices have very poor repeatability for the standard samples. Can we have a look at only these?
#key sensitive
```

