---
title: "DWARF Fluorescence PARAFAC"
author: "Cathrine Brecke Gundersen (NIVA)"
date: "3/23/2022"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r load required packages, include=FALSE}
Packages <- c("staRdom","dplyr", "tidyr")
lapply(Packages, library, character.only = TRUE)
library(htmltools)
```

### Introduction
- first attempt to PARAFAC the fluorescence EEM from the DWARF project 
- Aim: can we distinguish anthropogenic from natural DOM?
- procedure following: https://cran.r-project.org/web/packages/staRdom/vignettes/PARAFAC_analysis_of_EEM.html and https://doi.org/10.1039/C3AY41160E

```{r Load and combine data, echo=FALSE}
#library(devtools)
#install_github("MatthiasPucher/staRdom")
#library(staRdom)

eem_list1 = eem_import_dir("C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/DWARF-PARAFAC/new/")

eem_list2 = eem_import_dir("C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/DWARF-PARAFAC/new/new2/new3/")
getwd()


eem_list3 <- c(eem_list1, eem_list2)

#summary(eem_list2)
class(eem_list1)
```

```{r View EEM data, message=FALSE}
eem_overview_plot(eem_list1, spp=9, contour = TRUE)
dwarf_sum <- summary(eem_list2)  

#following to find samples to keep
write.csv(dwarf_sum,"C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/DWARF_summary_ALL23.csv", row.names = FALSE)
```
### Step 1. Upload data and exclude samples that do not belong to the project
- The following samples were excluded for not being part of the DWARF project
- Samples included from Otava and Karhov
```{r exclude samples that dont belong to the dwarf project, message=FALSE}
#The noisy range below 250 nm excitation and above 580 nm emission can be removed 
exclude <- list("sample" = c("v2021135", "v2021136","v2021137","v2021138","v2021139","v2021140","v2021141","v2021142","v2021143","v2021144","v2021145","v2021146","v2021147","v2021148","v2021149","v2021150","v2022117","v2022118","v2022119","v2022120","v2022121","v2022122","v2022123","v2022124","v2022125","v2022126","v2022127","v2022128","v2022129","v2022130","v2022153", "v2022154","v2022155","v2022156","v2022157","v2022158","v2022159","v2022160","v2022161","v2022164","v2022165","v20212837","v20212838","v20212839","v20212840","v20212841","v20212842","v20212843","v20212844","v20212845","v20212846","v20212847","v20213112","v20213113","v20213114","v20213142","v20213143","v20213144","v20213223","v20213224","v20213225","v20213226","v20213231","v20213232","v20213234","v20213276","v20213277","v20213278","v20213279","v20213280","v20213281","v20213371","v20213372","v20213373","v020213141NF","v020213142NF","v020213143NF","v020213144NF","v020213222NF","v020213223NF","v020213224NF","v020213225NF","v20213226NF","v20213233NF","v20213234NF","v20213275NF","v20213276NF","v20213277NF","v20213278NF","v20213279NF","v20213280NF","v20213281NF","v20213370NF","v20213371NF","v20213372NF","v20213373NF","v20213141NF","v20213142NF","v20213143NF","v20213144NF","v20213222NF","v20213223NF","v20213224NF","v20213225NF"))

#print(exclude)
print(exclude)
# exclude outliers if neccessary. if so, restart analysis
eem_list_ex <- eem_exclude(eem_list1, exclude)
```
### Step 2. Summaries the data
- EEM plots
- summary of pre-processing. Looks like samples have not been blank corrected, but Petr has confirmed. 
```{r Sample summary, echo=FALSE, message=FALSE}
eem_list2 <- eem_list_ex %>% eem_range(ex = c(250,Inf), em = c(300,580))
eem_overview_plot(eem_list2, spp=9, contour = TRUE)
knitr::kable(summary(eem_list2))
```
### Step 3. First PARAFAC attempt. Lower "resolution"
Aim: obtain the correct number of components
- creates and compares three different models with 2, 3, and 4 components each, respectively.

- the fit of the models is good (r2), very small increase going from 3 to 4 components
- the excitation and emission bands do not look perfect, worsen for every added component
- strong correlation between the components. Therefore, data is normalized. Does not help a lot though... 
- Interestingly, the third component of model 2 looks identical to 1000Lakes assumed contamination
```{r First model attempt, echo=FALSE}
 #10) First attempt PARAFAC model
# minimum and maximum of numbers of components
# parameter settings from stardom guideline
cores <- detectCores(logical = FALSE)
dim_min <- 2
dim_max <- 4

nstart <- 25 # number of similar models from which best is chosen
maxit = 5000 # maximum number of iterations in PARAFAC analysis
ctol <- 10^-6 # tolerance in PARAFAC analysis

# 11) calculating PARAFAC models, one for each number of components
#using non-negative constraints
pf1 <- eem_parafac(eem_list2, comps = seq(dim_min,dim_max), normalise = FALSE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)

# 12) rescale B and C modes to a maximum fluorescence of 1 for each component
pf1n <- lapply(pf1, eempf_rescaleBC, newscale = "Fmax")
#Compare the different models in plot
eempf_compare(pf1n, contour = TRUE)


pf4 <- eem_parafac(eem_list2, comps = 6, normalise = TRUE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, output = "all", cores = cores)

ssccheck <- eempf_ssccheck(pf4[[1]]$models, best = 3, cores = cores) # best 3 models are shown

eempf_plot_ssccheck(ssccheck)
```

```{r First model checking: correlation between components}
# 13) check for correlation between the components
#The number indicates which model to look at
#print(eempf_cortable(pf1n[[2]], normalisation=FALSE))
eempf_corplot(pf1n[[2]], progress = FALSE, normalisation = FALSE)

# 14) Normalise to reduce correlation between components
pf1x <- eem_parafac(eem_list2, comps = seq(dim_min,dim_max), normalise = TRUE, const = c("nonneg", "nonneg", "nonneg"), maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
# rescale B and C modes
pf1xn <- lapply(pf1x, eempf_rescaleBC, newscale = "Fmax")
#eempf_compare(pf1xn, contour = TRUE, normalisation = FALSE)

#Set normalisation TRUE if you want to see the actual data
#Set normalisation FALSE if you want to see the normalised data
#eempf_cortable(pf1xn[[2]], normalisation = FALSE)
eempf_corplot(pf1xn[[2]], progress = FALSE, normalisation = FALSE)
#eempf_compare(pf1xn, contour = TRUE, normalisation=FALSE) 
```

### Step 4. Evaluate extreme samples/wavelengths
- sometimes, extreme samples or emission/excitation wavelengths can be excluded from the sample set prior to making the PARAFAC model
- dont appear to be any strong outliers (value from 0 to 1)
- decides to keep all samples and wavelengths 
```{r Look for outliers in leverages}
# 15) Find and exclude outlier leverages
cpl <- eempf_leverage(pf1xn[[2]])
# plot leverage (nice plot)
eempf_leverage_plot(cpl,qlabel=0.1)

# samples, excitation and emission wavelengths to exclude, makes sense after calculation of leverage
#exclude1 <- list("ex" = c("250", "280"),
#                "em" = c(),
#                "sample" = c()
#)

# exclude outliers if neccessary. if so, restart analysis
#eem_list_ex1 <- eem_exclude(eem_list2, exclude1)

# plot leverage, not so nice plot but interactive to select what to exclude
# saved in exclude, can be used to start over again with eem_list_ex <- eem_list %>% eem_exclude(exclude) above
#exclude <- eempf_leverage_ident(cpl,qlabel=0.1)


# DID NOT MANAGE THE FOLLOWING. SOMETHING TO DO WITH SAMPLE NAMES?
#eempf_residuals_plot(pf1xn[[2]], eem_list2, residuals_only = TRUE, 
 #                    select = c("v20212857", "v20212836"), spp=6, cores = cores, contour = TRUE)
```


```{r Create new model without outliers}
#pf2 <- eem_parafac(eem_list_ex1, comps = seq(dim_min,dim_max), normalise = TRUE, maxit = maxit, nstart = nstart, ctol = ctol, cores = cores)
#pf2n <- lapply(pf2, eempf_rescaleBC, newscale = "Fmax")

#eempf_compare(pf2n, contour = TRUE, normalisation = FALSE)
#eempf_leverage_plot(eempf_leverage(pf2n[[2]]),qlabel=0.1)
```
### Step 5. Re-run PARAFAC model with increased "resolution"
- selects model no2 with 3 components
```{r PARAFAC increased}
# 16) ruN MODEL AGAIN WITH INCREASED
# minimum and maximum of numbers of components
#nstart <- 50 # number of similar models from which best is chosen
#maxit = 10000 # maximum number of iterations in PARAFAC analysis
#ctol <- 10^-10 # tolerance in PARAFAC analysis

#here we specify the number of components and so there is only one model
#pf3c <- eem_parafac(eem_list2, comps = 3, const = c("nonneg", "nonneg", "nonneg"), normalise = TRUE, maxit = maxit, nstart = nstart, output = "all", ctol = ctol, cores = cores)
#pf3cn <- lapply(pf3c, eempf_rescaleBC, newscale = "Fmax")
#eempf_compare(pf3cn, contour = TRUE)

# Save an object to a file
#saveRDS(pf3cn, file = "220406_DWARF_model.RData")
# Restore the object
pf3cn <- readRDS(file = "220406_DWARF_model.RData")

```

### Step 6. Model checking
- several options are available, with their different advantagous and disadvantagous
- some subjectivity is involved
```{r Model checking}
 # 17) Check the convergence behavior of the created models:
eempf_convergence(pf3cn[[1]])
#eempf_compare(pf3cn, contour=TRUE)
eempf_leverage_plot(eempf_leverage(pf3cn[[1]]))
#eempf_corplot(pf3cn[[1]], progress = FALSE)

#plot the resulting components and loadings
eempf_comp_load_plot(pf3cn[[1]], contour = TRUE)

# plot components in each sample, residual and whole sample
#WHen using this plot, perhaps choose only some samples at the time? Did not work now with all
#eempf_residuals_plot(pf3cn[[1]], eem_list2, cores = cores, contour = TRUE)

#calculate split_half analysis
sh <- splithalf(eem_list2, 3, normalise = TRUE, rand = FALSE, cores = cores, nstart = nstart, strictly_converging = TRUE, maxit = maxit, ctol = ctol)
splithalf_plot(sh)
sh %>% unlist(recursive = FALSE) %>% eempf_compare()

# 20) Tucker’s Congruency Coefficients is a value for the similarity of the splits
# 1 would be perfect similarity
tcc_sh_table <- splithalf_tcc(sh)
tcc_sh_table

sh_r <- splithalf(eem_list2, 3, normalise = TRUE, rand = TRUE, cores = cores, nstart = nstart, maxit = maxit, ctol = ctol)
splithalf_plot(sh_r)
tcc_sh_table2 <- splithalf_tcc(sh_r)
tcc_sh_table2

varimp <- eempf_varimp(pf3cn[[1]], eem_list2, cores = cores)
varimp
```

### Step 6. RESULTS: Created model is exported to OpenFluor to compare with other studies and identify the components
From OpenFluor the following can be linked to the model components: 
* C1: Humic-material, larger-sized, characteristic of soil, sediment and freshwater environments, recalcitrant soil or plant material, highly resistant to microbial degradation but suspictible to photodecay (https://doi.org/10.1016/j.geoderma.2017.06.029), similar to soil-fulvic material (10.1080/00785236.2003.10409512), larger sized humics from cultivated soils (10.1016/j.jaridenv.2019.04.013),
Terrestrial humic-like fluorescence in high nutrient and wastewater impacted environments (10.1021/es103015e), Humic-like
Terrestrial delivered OM (10.1016/j.watres.2014.01.053)
* C2: Humic-material, mid-sized, characteristic of soil, sediment and freshwater environments, recalcitrant soil or plant material, highly resistant to microbial degradation but susceptible to photodecay (https://doi.org/10.1016/j.geoderma.2017.06.029), terrestrial humic-like (e.g. Painter et al.2018); microbially- or photo-chemically altered organic matter (Yamashita et al. 2013) (10.1088/1748-9326/abac36)
* C3: protein-like material (tryptophane/tyrosine), often linked to anthropogenic distrubances (10.1890/12-0825.1), protein-like (10.1890/12-0825.1), biological- and fresh production (10.1016/j.marchem.2019.103720)


```{r Figure, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
#tinytex::reinstall_tinytex()
tabl <- "
| Component | Ex. max (nm)  | Em. max (nm)  |
|-----------|:-------------:|--------------:|
|C1         | 265/365       |         485   |
|C2         | 250/310       |         412   |
|C3         | 280           |         321   |
"
cat(tabl)

```


```{r Model Export, eval = FALSE}
#Export to OpenFluor
eempf_openfluor(pf3cn[[1]], file = "22_CBG_DWARF_3comps_2.txt")
#Export model for use
eempf_export(pf3cn[[1]], export = "C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/220401_PARAFAC_DWARF_3comp.txt", Fmax = FALSE)


#Make report
eempf_report(pf5n[[1]], export = "parafac_report.html", eem_list = eem_list_ex, shmodel = sh, performance = TRUE)
#eempf_report(pf3[[1]], export = "parafac_report.html")

#Reverse normalisation
p4 <- norm2A(pf5[[1]])

#Export model
#No difference in output from Fmax TRUE or FALSE for pf5m(1)
#p4 fra pf5n ga samma output som over
#Now they represent two different outputs, BUT they are relatively the same, the variation among the samples
eempf_export(pf_2cn[[1]], export = "C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/PARAFAC_DWARF_2comp.txt", Fmax = FALSE)
eempf_export(pf_3cn[[1]], export = "C:/Users/CBG/OneDrive - NIVA/1 Projects/DWARF/PARAFAC_DWARF_3comp.txt", Fmax = FALSE)
#the fluorescence indices have very poor repeatability for the standard samples. Can we have a look at only these?
#key sensitive
```

